---
title: "CFSS Homework 03"
author: "Cory J. Cascalheira"
date: "December 21, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1
___
Load dependencies.

```{r message=FALSE}
library(tidyverse)
library(readxl)
```

Set the default theme for `ggplot2`.

```{r}
theme_set(theme_minimal())
```

Load the data.

```{r}
load("J:/sample_data/dadmom.rda")
dadmom
```

Tidy the data frame.

```{r}
# Tidy parents and names
(parents <- dadmom %>%
  # Remove income columns
  select(-starts_with("inc")) %>%
  # Gather except family ID
  gather(key = parent, value = name, -famid) %>%
  # Make parent variable make more sense
  mutate(parent = fct_recode(factor(parent), "father" = "named", "mother" = "namem"))
)

# Tidy the income, make sure order matches parents
(income <- dadmom %>%
  # Remove name columns
  select(-starts_with("name")) %>%
  # Gather except family ID
  gather(key = parent, value = income, -famid) %>%
  # Drop family ID and parent columns
  select(income)
)

# Bind columns
(dadmom_new <- bind_cols(parents, income)
)
```

# Part 2
___
Load the data.

```{r}
library(gapminder)
hiv <- read_excel("J:/sample_data/hiv-prevalence.xlsx")
hiv
```

Fix the country column.

```{r}
colnames(hiv)[1] <- "country"
```

Are the same countries in both datasets?

```{r}
# Which countries are in the gapminder dataset
gapminder_names <- unique(as.character(gapminder$country))

# Filter the HIV dataset to include only those countries in gapminder
(hiv2 <- hiv %>%
  filter(country %in% gapminder_names)
)
```

Tidy the HIV data frame.

```{r}
(hiv3 <- hiv2 %>%
  gather(key = year, value = hiv_prevalence, -country)
)
```

Transform year variable into an integer and HIV prevalence into a numeric.

```{r}
hiv3$year <- as.integer(hiv3$year)
hiv3$hiv_prevalence <- as.numeric(hiv3$hiv_prevalence)

# Verify change
head(hiv3)
```

## HIV Prevalence & Life Expectancy

I will use a `left_join()` to keep all original columns and observations from `gapminder`, effectively appending the HIV prevalence variable to the dataset.

```{r}
(gap_hiv <- left_join(gapminder, hiv3, by = c("country", "year")))
```

What is the relationship between HIV prevalence and life expectancy?

```{r}
# Filter out missing values
gap_hiv_complete <- gap_hiv %>%
  filter(!is.na(hiv_prevalence))

# Simple linear model
lm(lifeExp ~ hiv_prevalence, data = gap_hiv_complete)
```

There is a negative relationship between life expectancy and HIV prevalence. For each 1% increase of infection in the adult population, the life expectancy decreases by an average of 1.42 years.

```{r fig.height=7, fig.width=7, message=FALSE}
gap_hiv_complete %>%
  # Add continent for more detail
  ggplot(aes(x = hiv_prevalence, y = lifeExp, color = continent)) +
  geom_point(size = 2, alpha = 0.5) +
  # Prevent color aesthetic from inheriting
  geom_smooth(aes(color = NULL), se = FALSE, color = "cadetblue", size = 2) +
  # Provide nice labels
  labs(title = "Life Expectancy Predicted by HIV Infection Rate Among Adults",
       caption = "Source: Gapminder",
       x = "HIV Prevalence",
       y = "Life Expectancy",
       color = "Continent") +
  # Add custom colors
  scale_color_brewer(type = "qual", palette = "Paired") +
  # Move legend
  theme(legend.position = "top")
```

The trend is negative and drops sharply.

## Continents with NA

```{r}
gap_hiv %>%
  group_by(continent) %>%
  # Count missing values
  count(is.na(hiv_prevalence)) %>%
  # Clean column name
  rename("missing" = "is.na(hiv_prevalence)") %>%
  # Filter out non-missing values
  filter(missing == TRUE) %>%
  # Bar chart of results
  ggplot(aes(x = fct_rev(fct_reorder(continent, n)), y = n)) +
    # Use columns instead of bars to plot precise number
    geom_col() +
    # Add labels
    labs(title = "Missing Data on HIV Prevalence by Continent",
         caption = "Source: Gapminder",
         x = NULL,
         y = "Number of Missing Cases")
```

Alarmingly, even though African countries have the most cases of HIV, the dataset is least complete on the African continent. The HIV epidemic is likely more widespread than the data suggest.

# Part 3
___
Load the two datasets.

```{r message=FALSE, warning=FALSE}
modern <- read_csv("J:/sample_data/SCDB_2017_01_justiceCentered_Citation.csv")
legacy <- read_csv("J:/sample_data/SCDB_Legacy_03_justiceCentered_Citation.csv")
```

What do the variables of interest mean?

* `caseIssuesID`: unique key identifier
* `term`: term when decision handed down; has changed over the years
* `justice`: identification number of justices
* `justiceName`: name of justice
* `decisionDirection`: numeric variable; 1 = conservative, 2 = liberal, 3 = unspecified
* `majVotes`: number of justices voting in the majority
* `minVotes`: number fo justices voting in the dissent
* `majority`: whether justice voted with 1 = dissent or 2 = majority 
* `chief`: chief justice of the court
* `dateDecision`: date that decision was delivered
* `decisionType`: type of court decision, with values of:
    1.	opinion of the court (orally argued)
    2.	per curiam (no oral argument)
    4.	decrees
    5.	equally divided vote
    6.	per curiam (orally argued)
    7.	judgment of the Court (orally argued)
    8.	seriatim

Select only these variables from both datasets.

```{r}
modern_sm <- modern %>%
  select(caseIssuesId, term, justice, justiceName, decisionDirection,
         majVotes, minVotes, majority, chief, dateDecision, decisionType)

legacy_sm <- legacy %>%
  select(caseIssuesId, term, justice, justiceName, decisionDirection,
         majVotes, minVotes, majority, chief, dateDecision, decisionType)
```

Bind the rows to create a combined data frame.

```{r}
(supreme <- bind_rows(modern_sm, legacy_sm))
```

## Percentage of Cases Decided by One-vote Margin

The observational unit is at the individual justice level per case. First, we need to count the number of cases per term. 

```{r}
(cases_per_term <- supreme %>%
  # n represents the number of justices
  count(term, caseIssuesId) %>%
  group_by(term) %>%
  summarize(
    num_cases = n()
  ))
```

Count the number of decisions with a one-vote margin per term.

```{r}
(num_marginal_votes <- supreme %>%
  # Create new margin variable
  mutate(margin = majVotes - minVotes) %>%
  # n represents the number of justices
  count(term, caseIssuesId, margin) %>%
  # Only cases with a one-vote margin
  filter(margin == 1) %>%
  # Group cases by term
  group_by(term) %>%
  summarize(
    num_marginal_votes = n()
  ))
```

Now, we can `left_join` the two data frames. Missing values for the variable `marginal_votes` indicates that the term had no one-vote margins.

```{r}
(one_vote_terms <- left_join(cases_per_term, num_marginal_votes, by = "term"))
```

Replace the missing values with the number zero. This zero represents an absence of marginal votes, not a margin of zero (which would represent a split vote).

```{r}
# Which rows are missing values?
(no_margin <- which(is.na(one_vote_terms$num_marginal_votes)))

# Subset data frame to replace missing values with 0
one_vote_terms$num_marginal_votes[no_margin] <- 0

# Verify that the process worked
head(one_vote_terms)
```

Percentage of cases decided by a one-vote margin.

```{r}
one_vote_terms %>%
  mutate(percent_marginal = num_marginal_votes / num_cases)
```

A graphical representation of the percentage of decisions involving one-vote margins per term.

```{r}
one_vote_terms %>%
  mutate(percent_marginal = num_marginal_votes / num_cases) %>%
#  filter(percent_marginal != 0) %>%
  ggplot(aes(x = term, y = percent_marginal)) +
    geom_col() +
    scale_y_continuous(labels = scales::percent) +
    labs(title = "Percent of Supreme Court Decisions Involving One-Vote Margins",
         subtitle = "1791 - 2016",
         x = "Term",
         y = NULL)
```

## Percentage of Cases Justice Antonin Scalia in Majority

