---
title: "CFSS Homework 03"
author: "Cory J. Cascalheira"
date: "December 21, 2018"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1
___
Load dependencies.

```{r message=FALSE}
library(tidyverse) # utility & visualization
library(readxl) # import excel data
library(lubridate) # convert to date
```

Set the default theme for `ggplot2`.

```{r}
theme_set(theme_minimal())
```

Load the data.

```{r}
load("J:/sample_data/dadmom.rda")
dadmom
```

Tidy the data frame.

```{r}
# Tidy parents and names
(parents <- dadmom %>%
  # Remove income columns
  select(-starts_with("inc")) %>%
  # Gather except family ID
  gather(key = parent, value = name, -famid) %>%
  # Make parent variable make more sense
  mutate(parent = fct_recode(factor(parent), "father" = "named", "mother" = "namem"))
)

# Tidy the income, make sure order matches parents
(income <- dadmom %>%
  # Remove name columns
  select(-starts_with("name")) %>%
  # Gather except family ID
  gather(key = parent, value = income, -famid) %>%
  # Drop family ID and parent columns
  select(income)
)

# Bind columns
(dadmom_new <- bind_cols(parents, income)
)
```

# Part 2
___
Load the data.

```{r}
library(gapminder)
hiv <- read_excel("J:/sample_data/hiv-prevalence.xlsx")
hiv
```

Fix the country column.

```{r}
colnames(hiv)[1] <- "country"
```

Are the same countries in both datasets?

```{r}
# Which countries are in the gapminder dataset
gapminder_names <- unique(as.character(gapminder$country))

# Filter the HIV dataset to include only those countries in gapminder
(hiv2 <- hiv %>%
  filter(country %in% gapminder_names)
)
```

Tidy the HIV data frame.

```{r}
(hiv3 <- hiv2 %>%
  gather(key = year, value = hiv_prevalence, -country)
)
```

Transform year variable into an integer and HIV prevalence into a numeric.

```{r}
hiv3$year <- as.integer(hiv3$year)
hiv3$hiv_prevalence <- as.numeric(hiv3$hiv_prevalence)

# Verify change
head(hiv3)
```

## HIV Prevalence & Life Expectancy

I will use a `left_join()` to keep all original columns and observations from `gapminder`, effectively appending the HIV prevalence variable to the dataset.

```{r}
(gap_hiv <- left_join(gapminder, hiv3, by = c("country", "year")))
```

What is the relationship between HIV prevalence and life expectancy?

```{r}
# Filter out missing values
gap_hiv_complete <- gap_hiv %>%
  filter(!is.na(hiv_prevalence))

# Simple linear model
lm(lifeExp ~ hiv_prevalence, data = gap_hiv_complete)
```

There is a negative relationship between life expectancy and HIV prevalence. For each 1% increase of infection in the adult population, the life expectancy decreases by an average of 1.42 years.

```{r fig.height=7, fig.width=7, message=FALSE}
gap_hiv_complete %>%
  # Add continent for more detail
  ggplot(aes(x = hiv_prevalence, y = lifeExp, color = continent)) +
  geom_point(size = 2, alpha = 0.5) +
  # Prevent color aesthetic from inheriting
  geom_smooth(aes(color = NULL), se = FALSE, color = "cadetblue", size = 2) +
  # Provide nice labels
  labs(title = "Life Expectancy Predicted by HIV Infection Rate Among Adults",
       caption = "Source: Gapminder",
       x = "HIV Prevalence",
       y = "Life Expectancy",
       color = "Continent") +
  # Add custom colors
  scale_color_brewer(type = "qual", palette = "Paired") +
  # Move legend
  theme(legend.position = "top")
```

The trend is negative and drops sharply.

## Continents with NA

```{r}
gap_hiv %>%
  group_by(continent) %>%
  # Count missing values
  count(is.na(hiv_prevalence)) %>%
  # Clean column name
  rename("missing" = "is.na(hiv_prevalence)") %>%
  # Filter out non-missing values
  filter(missing == TRUE) %>%
  # Bar chart of results
  ggplot(aes(x = fct_rev(fct_reorder(continent, n)), y = n)) +
    # Use columns instead of bars to plot precise number
    geom_col() +
    # Add labels
    labs(title = "Missing Data on HIV Prevalence by Continent",
         caption = "Source: Gapminder",
         x = NULL,
         y = "Number of Missing Cases")
```

Alarmingly, even though African countries have the most cases of HIV, the dataset is least complete on the African continent. The HIV epidemic is likely more widespread than the data suggest.

# Part 3
___
Load the two datasets.

```{r message=FALSE, warning=FALSE}
modern <- read_csv("J:/sample_data/SCDB_2017_01_justiceCentered_Citation.csv")
legacy <- read_csv("J:/sample_data/SCDB_Legacy_03_justiceCentered_Citation.csv")
```

What do the variables of interest mean?

* `caseIssuesID`: unique key identifier
* `term`: term when decision handed down; has changed over the years
* `justice`: identification number of justices
* `justiceName`: name of justice
* `decisionDirection`: numeric variable; 1 = conservative, 2 = liberal, 3 = unspecified
* `majVotes`: number of justices voting in the majority
* `minVotes`: number fo justices voting in the dissent
* `majority`: whether justice voted with 1 = dissent or 2 = majority 
* `chief`: chief justice of the court
* `dateDecision`: date that decision was delivered
* `decisionType`: type of court decision, with values of:
    1.	opinion of the court (orally argued)
    2.	per curiam (no oral argument)
    4.	decrees
    5.	equally divided vote
    6.	per curiam (orally argued)
    7.	judgment of the Court (orally argued)
    8.	seriatim

Select only these variables from both datasets.

```{r}
modern_sm <- modern %>%
  select(caseIssuesId, term, justice, justiceName, decisionDirection,
         majVotes, minVotes, majority, chief, dateDecision, decisionType)

legacy_sm <- legacy %>%
  select(caseIssuesId, term, justice, justiceName, decisionDirection,
         majVotes, minVotes, majority, chief, dateDecision, decisionType)
```

Bind the rows to create a combined data frame.

```{r}
(supreme <- bind_rows(modern_sm, legacy_sm))
```

There was no issue binding the two data frames together. This might be due to the selection of columns prior to binding, or the fact that I downloaded the datasets directly from the Supreme Court Database.

## Percentage of Cases Decided by One-vote Margin

The observational unit is at the individual justice level per case. First, we need to count the number of cases per term. 

```{r}
(cases_per_term <- supreme %>%
  # n represents the number of justices
  count(term, caseIssuesId) %>%
  group_by(term) %>%
  summarize(
    num_cases = n()
  ))
```

Count the number of decisions with a one-vote margin per term.

```{r}
(num_marginal_votes <- supreme %>%
  # Create new margin variable
  mutate(margin = majVotes - minVotes) %>%
  # n represents the number of justices
  count(term, caseIssuesId, margin) %>%
  # Only cases with a one-vote margin
  filter(margin == 1) %>%
  # Group cases by term
  group_by(term) %>%
  summarize(
    num_marginal_votes = n()
  ))
```

Now, we can `left_join` the two data frames. Missing values for the variable `marginal_votes` indicates that the term had no one-vote margins.

```{r}
(one_vote_terms <- left_join(cases_per_term, num_marginal_votes, by = "term"))
```

Replace the missing values with the number zero. This zero represents an absence of marginal votes, not a margin of zero (which would represent a split vote).

```{r}
# Which rows are missing values?
(no_margin <- which(is.na(one_vote_terms$num_marginal_votes)))

# Subset data frame to replace missing values with 0
one_vote_terms$num_marginal_votes[no_margin] <- 0

# Verify that the process worked
head(one_vote_terms)
```

Percentage of cases decided by a one-vote margin.

```{r}
one_vote_terms %>%
  mutate(percent_marginal = num_marginal_votes / num_cases)
```

A graphical representation of the percentage of decisions involving one-vote margins per term.

```{r}
one_vote_terms %>%
  mutate(percent_marginal = num_marginal_votes / num_cases) %>%
#  filter(percent_marginal != 0) %>%
  ggplot(aes(x = term, y = percent_marginal)) +
    geom_col() +
    # Percent added to values on y-axis and force origin to start at 0
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    # Customize breaks along x-axis
    scale_x_continuous(breaks = seq(1790, 2016, by = 25)) +
    labs(title = "Percent of Supreme Court Decisions Involving One-Vote Margins",
         subtitle = "1791 - 2016",
         x = "Term",
         y = NULL,
         caption = "Source: Supreme Court Database")
```

## Percentage of Cases Justice Antonin Scalia in Majority

Remember, if `majority == 2`, then Justice Scalia was in the majority.

```{r}
# Total number of cases for Scalia
(scalia_total <- supreme %>%
  filter(justiceName == "AScalia") %>%
 count(term))

# Cases in which Scalia was in majority
(scalia_majority <- supreme %>%
  filter(justiceName == "AScalia") %>%
  group_by(term, majority) %>%
  summarize(num_majority = n()) %>%
  filter(majority == 2))

# Join data frames
left_join(scalia_total, scalia_majority, by = "term") %>%
  group_by(term, n, num_majority) %>%
  summarize(percent_majority = num_majority / n) %>%
  ggplot(aes(x = term, y = percent_majority)) +
    geom_point() +
    geom_line(color = "cadetblue") +
    # Scales adds percent sign to values on y-axis
    scale_y_continuous(labels = scales::percent, limits = c(.5, 1)) +
    scale_x_continuous(breaks = seq(1986, 2016, by = 4)) +
    labs(title = "Justice Scalia in Majority",
         subtitle = "Percent of Cases per Term Over His Career",
         x = "Term",
         y = NULL,
         caption = "Source: Supreme Court Database")
```

For most of his career, Justice Scalia voted with the majority opinion on more than 70% of the cases for which he delivered judgement.

## Percentage of Cases Non-Unanimous

We already calculated the number cases per term. This aggregate data frame will be useful in a moment.

```{r}
head(cases_per_term)
```

How many cases were non-unanimous in each term? As long as there is at least one dissent, then the decision was not unanimous. That is, the case is non-unanimous if `minVotes > 0`.

```{r}
(num_non_unanimous <- supreme %>%
  # Group all justices into each case; n represents number of justices
  count(caseIssuesId, term, minVotes) %>%
  # Only non-unanimous decisions
  filter(minVotes > 0) %>%
  group_by(term) %>%
  # Number of non-unanimous decisions per term
  summarize(
    non_unanimous_decisions = n()
  ))
```

Now we can merge with the data frame of the sum of cases per term.

```{r}
head(dissent <- left_join(cases_per_term, num_non_unanimous, by = "term"))
```

Which observations contain missing values?

```{r}
(no_dissent <- which(is.na(dissent$non_unanimous_decisions)))
```

Replace those missing values with zero. Zero indicates that all cases for that term were decided unanimously.

```{r}
dissent$non_unanimous_decisions[no_dissent] <- 0

# Verify that process worked
head(dissent)
```

Visualize the percentage of cases per term that were non-unanimous.

```{r}
dissent %>%
  # Calculate percentage of non-unanimous cases
  mutate(percent_non_unanimous = non_unanimous_decisions / num_cases) %>%
  ggplot(aes(x = term, y = percent_non_unanimous)) +
    geom_col() +
    # Add percent sign, extend y-axis to 100%, customize breaks, force origin to start at 0
    scale_y_continuous(labels = scales::percent, limits = c(0, 1), 
                       breaks = seq(0, 1, 0.1), expand = c(0, 0)) +
    # Customize breaks along x-axis
    scale_x_continuous(breaks = seq(1790, 2016, by = 25)) +
    labs(title = "Non-Unanimity of Supreme Court Decisions",
         subtitle = "Percent of Cases Not Unanimous per Term",
         x = "Term",
         y = NULL,
         caption = "Source: Supreme Court Database")
```

Aside from an outlier in the early 1790s, the graph demonstrates that the non-unanimity of court decisions has increased over time. Has the modern era involved more divisive cases?

## Percentage of Cases in Conservative Direction

Again, we can use our data frame of the sum of cases per term as the denominator of the ratio. 

```{r}
head(cases_per_term)
```

First, we need to calculate the number of cases decided in the conservative direction. Remember, if `decisionDirection == 1`, then the case was decided in the conservative direction.

```{r}
(num_conservative <- supreme %>%
  # Lump individual justices into single cases
  count(term, caseIssuesId, decisionDirection) %>%
  # Keep only the cases where the decision was conservative
  filter(decisionDirection == 1) %>%
  # Group the cases by term
  group_by(term) %>%
  # Number of cases that were conservatively decided
  summarize(
    conservative_decisions = n()
  ))
```

Join the two data frames and calculate the percentage of conservative decisions.

```{r}
(conservative <- left_join(cases_per_term, num_conservative, by = "term") %>%
  # Percentage of conservative decisions per term
  mutate(percent_conservative = conservative_decisions / num_cases))
```

Visualize the percentage of decisions in the conservative direction over time.

```{r fig.width=10}
conservative %>%
  ggplot(aes(x = term, y = percent_conservative)) +
    geom_point() +
    # Show trend line
    geom_line(color = "cadetblue") +
    # Add percent signs, breaks, expand the y-axis, and start origin at 0
    scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.1),
                       limits = c(0, 1), expand = c(0, 0)) +
    # Customize breaks along x-axis
    scale_x_continuous(breaks = seq(1790, 2016, by = 25)) +
    # Add horizontal line at 50% for comparison purposes
    geom_hline(yintercept = 0.5, color = "blue") +
    labs(title = "Has the Supreme Court of the U.S. Become More Conservative?",
         subtitle = "Percent of Decisions in the Conservative Direction, 1791 - 2016",
         x = "Term",
         y = NULL, 
         caption = "Source: Supreme Court Database")
```

## Decisions per Month

Start by transforming the `dateDecision` variable into a date.

```{r}
supreme$dateDecision <- mdy(supreme$dateDecision)

# Did the transformation work?
class(supreme$dateDecision)
```

Create a new column for the month that the decision was delivered, then group the cases by term and month to count the decisions per month. For fun, we can answer the question visually in one long pipeline.

```{r fig.height=10, fig.width=10}
supreme %>%
  # Create the month variable
  mutate(month = month(dateDecision)) %>%
  # Condense the justices into one case, pass along month column
  count(term, caseIssuesId, month) %>%
  # Group by the months per term
  group_by(term, month) %>%
  # Count the number of decisions delivered in a particular month
  summarize(
    num_decisions = n()
  ) %>%
  ggplot(aes(x = term, y = num_decisions)) +
    geom_point() +
    facet_wrap(~ fct_recode(factor(month),
                            "January" = "1",
                            "February" = "2",
                            "March" = "3",
                            "April" = "4",
                            "May" = "5",
                            "June" = "6",
                            "July" = "7",
                            "August" = "8",
                            "September" = "9",
                            "October" = "10",
                            "November" = "11",
                            "December" = "12"),
               ncol = 3) +
    labs(title = "Number of Supreme Court Decisions per Month",
         subtitle = "1791-2016",
         x = "Term",
         y = NULL,
         caption = "Source: Supreme Court Database") +
    # Change default text size
    theme(text = element_text(size = 16))
```

The Supreme Court delivers most of its decisions in January, February, and March. This could be related to the start of each term. Few decisions occur during July, and hardly any during August and September. Are these months used for deliberation?